---
title: "Notes for session 1"
author: "Ege Rubak"
date: "August 2022"
output: github_document
---

## Introduction

```{r, echo=FALSE, message=FALSE}
library(spatstat)
```

_Spatial data_= data attributed to spatial locations

Three main types of spatial data:

* _spatial variable ("field")_, eg temperature
* _regional aggregate data_, eg accident counts in each state
* _spatial point patterns_, eg locations of crimes/accidents

```{r, echo=FALSE, results="hide", fig.width=9, fig.height=3, out.width="100%"}
W <- Window(austates)
f <- function(x,y){ 50+y }
X <- rpoint(70, f, win=W)
Z <- scaletointerval(density(X, 3), from=10, to=40)
agg <- quadratcount(X, tess=austates)
plot(solist(field=Z, regional=agg, points=X), main="", box=FALSE)
```

This workshop is about the analysis of _spatial point patterns_

We will use the `spatstat` package in `R`

```{r}
library(spatstat)
```

## Spatial point pattern terminology

### Points

The "points" in a point pattern are the spatial locations
where the events or objects were observed. They are specified
by spatial coordinates. **NOTE:** In all that follows and for all functions in
`spatstat` the coordinates are assumed to be **projected coordinates in Euclidean
space**. Do not analyse geographic coordinates (latitude and longitude) directly
in `spatstat` -- project them first!

```{r, echo=FALSE, results="hide"}
xx <- c(0.3, 0.5, 0.7)
yy <- c(0.3, 0.7, 0.5)
XX <- ppp(xx, yy, 0:1, 0:1)
LL <- psp(c(xx[1], xx[1], xx[2], xx[2], xx[3], xx[3]),
          c(0,     yy[1], 0,     yy[2], 0,     yy[3]),
	  c(xx[1], 0,     xx[2], 0,     xx[3], 0),
	  c(yy[1], yy[1], yy[2], yy[2], yy[3], yy[3]),
	  owin())
plot(LL, main="", lwd=2, lty=2)
plot(XX, add=TRUE, pch=16, cex=1.5)
```

### Window

The window $W$ is the spatial region where
points were (or could have been) observed.

```{r, echo=FALSE, results="hide"}
plot(solist(unmark(chorley), Window(chorley)), main="", main.panel="")
```

### Covariates

Covariates are explanatory variables (which might "explain"
any spatial variation in the abundance of points, for example).

Many covariates take the form of a function $Z(u), \quad u \in W$
defined at every spatial location $u$.

```{r, echo=FALSE, results="hide"}
plot(Z, main="",box=FALSE)
```

Alternatively, other kinds of spatial data can be treated as
explanatory data. Usually we need to translate them into spatial functions
for use in analysis.

### Marks

Marks are attributes of the individual events or things.

In a spatial point pattern of trees, the trees might be classified
into different species, and each tree carries a mark ("label")
indicating which species it belongs to.

```{r, echo=FALSE, results="hide"}
plot(urkiola, main="", cols=2:3)
```

Marks are methodologically different from covariates:
marks are part of the "response", not the "
explanatory variable"

## Software and data 

### Spatstat

```{r}
library(spatstat)
```

A point pattern dataset is represented an object belonging to the class
`"ppp"` (planar point pattern). Some datasets are included in the package:

```{r}
gordon
class(gordon)
```

```{r}
plot(gordon)
```

The spatial coordinates of the points can be extracted by
`as.data.frame`:

```{r}
head(as.data.frame(gordon))
```

The window of observation for a point pattern can be extracted by (notice the upper case W in `Window` -- it is important):

```{r}
W <- Window(gordon)
W
class(W)
```

This is an object of class `"owin"` (observation window)
representing a spatial region.

If the points also carry _marks_, the marks can be extracted by
`marks()` or `as.data.frame()`:

```{r}
hyytiala
marks(hyytiala)
```

If the marks are a `factor` (categorical variable) then this specifies
a classification of the points into different groups.

The marks could also be numeric:

```{r}
longleaf
marks(longleaf)
```

The marks could be multivariate:

```{r}
finpines
head(marks(finpines))
```

Other kinds of spatial objects in `spatstat` include:

* pixel images: class `"im"`
* spatial patterns of line segments: class `"psp"`
* spatial tessellations: class `"tess"`

### Wrangling data

In this workshop, we will use datasets which are already
installed in _spatstat_, because time is short.

In practice, you would need to import your own data into `R`.

Data can be provided in many different file formats

* text file, CSV file
* shapefile
* `netcdf` file

The `spatstat` package does not support reading and writing of
files in different formats. This would be poor software design.

Instead, if you need to read files in a particular format,
we recommend that you find an appropriate `R` package
which is designed to read and write
that specific file format. Once the data have been read into `R`, then
you can use _another_ `R` package to convert the data into
objects recognised by `spatstat`. 

It is often enough to use 
the functions `read.table` and `read.csv` in the base `R` system
which will read simple text files containing columns of data and
store them in `R` as a `data.frame`. 

For full details please read the
[free copy of Chapter 3 the spatstat book](http://book.spatstat.org/sample-chapters/chapter03.pdf)
