---
title: 'Lab 2: Poisson models'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = FALSE)
```

```{r, include=FALSE}
library(spatstat)
```

### Exercise 1

The command `rpoispp(100)` generates realisations of the
Poisson process with intensity $\lambda = 100$ in the unit square.

1.  Repeat the command `plot(rpoispp(100))` several times
    to build your intuition about the appearance of a completely
    random pattern of points.

2.  Try the same thing with intensity $\lambda = 1.5$.


### Exercise 2

Returning to the Japanese Pines data,

1.  Fit the uniform Poisson point process model to the Japanese
    Pines data
    ```{r}
    ppm(japanesepines~1)
    ```

2.  Read off the fitted intensity. Check that this is the correct
    value of the maximum likelihood estimate of the intensity.

### Exercise 3

The `japanesepines` dataset is believed to exhibit
spatial inhomogeneity.

1.  Plot a kernel smoothed intensity estimate.

2.  Fit the Poisson point process models with loglinear intensity
    (trend formula `~x+y`) and log-quadratic intensity (trend
    formula `~polynom(x,y,2)`) to the Japanese Pines data.

3.  extract the fitted coefficients for these models using
    `coef`.

4.  Plot the fitted model intensity (using `plot(fit)`)

5.  perform the Likelihood Ratio Test for the null hypothesis of a
    loglinear intensity against the alternative of a log-quadratic
    intensity, using `anova`.

6.  Generate 10 simulated realisations of the fitted log-quadratic
    model, and plot them, using `plot(simulate(fit, nsim=10))` where 
    `fit` is the fitted model.

### Exercise 4

The `bei` dataset gives the locations of trees in a survey area
with additional covariate information in a list `bei.extra`.

1.  Fit a Poisson point process model to the data which assumes
    that the intensity is a loglinear function of terrain slope
    and elevation (hint: use `data = bei.extra` in `ppm`).

2.  Read off the fitted coefficients and write down the fitted
    intensity function.

3.  Plot the fitted intensity as a colour image.

4.  extract the estimated variance-covariance matrix of the
    coefficient estimates, using `vcov`.

5.  Compute and plot the standard error of the intensity estimate
    (see `help(predict.ppm)`).

### Exercise 5

The dataset `ants` is a multitype point pattern representing the
locations of nests of two species of ants.

1.  plot the data.

2.  Fit the model `ppm(ants ~ marks)` and interpret the result.
    Compare the result with `summary(ants)` and explain
    the similarities.

3.  Fit the model `ppm(ants ~ marks + x)` and write down an
    expression for the fitted intensity function.

4.  Fit the model `ppm(ants ~ marks * x)` and write down an
    expression for the fitted intensity function.

5.  Compute the fitted intensities of the three models fitted above 
    using `predict` and plot the results.

6.  Explain the difference between the models fitted by
    `ppm(ants ~ marks + x)` and `ppm(ants ~ marks * x)` .

## Exercise 6

The study region for the ants’ nests data `ants` is
divided into areas of "scrub" and "field". We want to fit a Poisson
model with different intensities in the field and scrub areas.

The coordinates of two points on the boundary line between field and
scrub are given in `ants.extra$fieldscrub`. First construct a function that
determines which side of the line we are on:

```{r}
fs <- function(x,y) {
  ends <- ants.extra$fieldscrub
  angle <- atan(diff(ends$y)/diff(ends$x))
  normal <- angle + pi/2
  project <- (x - ends$x[1]) * cos(normal) + (y - ends$y[1]) * sin(normal)
  factor(ifelse(project > 0, "scrub", "field"))
}
```

Now fit the models:
```{r, eval=FALSE}
ppm(ants ~ marks + side, data = list(side=fs))
ppm(ants ~ marks * side, data = list(side=fs))
```
and interpret the results.

### Exercise 7

The `update` command can be used to re-fit a point
process model using a different model formula.

1.  Type the following commands and interpret the results:
    ```{r}
    fit0 <- ppm(japanesepines ~ 1)
    fit1 <- update(fit0, . ~ x)
    fit1
    fit2 <- update(fit1, . ~ . + y)
    fit2
    ```

2.  Now type `step(fit2)` and interpret the results.
    

### Exercise 8

Fit Poisson point process models to the Japanese Pines data, with
the following trend formulas. Read off an expression for the fitted
intensity function in each case.

|Trend formula     |Fitted intensity function                   |
|:-----------------|:-------------------------------------------|
|`~1`              |                                            |
|`~x`              |                                            |
|`~sin(x)`         |                                            |
|`~x+y`            |                                            |
|`~polynom(x,y,2)` |                                            |
|`~factor(x < 0.4)`|                                            |

### Exercise 9

Make image plots of the fitted intensities for the inhomogeneous
models above.
