---
title: 'Lab 3: Dependence -- solution'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

```{r, message=FALSE}
library(spatstat)
```

### Exercise 1

The `swedishpines` dataset was recorded in a study plot
in a large forest. We shall assume the pattern is stationary.

1.  Calculate the estimate of the $K$-function using `Kest`.

    The estimation is done with:
    
    ```{r}
    K <- Kest(swedishpines)
    ```

2.  Plot the estimate of $K(r)$ against $r$

    To plot the K-function, we do:
    
    ```{r, results='hide'}
    plot(K, main = "K-function")
    ```

3.  Plot the estimate of $K(r) - \pi\!r^2$ against $r$ (Hint: look at the `fmla`
    argument in `plot.fv`).

    The estimated K-function subtracted $\pi\!r^2$ can be done via the `fmla`
    (formula) interface:

    ```{r}
    plot(K, . - pi*r^2 ~ r, main = "Normalized K-function",
         legendpos = "bottomright")
    ```
    
4.  Calculate and plot an estimate of the pair correlation function
    using `pcf`.
  
    The pair-correlation is also compute straight-forwardly:
    
    ```{r}
    pcorf <- pcf(swedishpines)
    plot(pcorf)
    ```

5.  Draw tentative conclusions from these plots about interpoint
    interaction in the data.

    Assuming a homogeneous point pattern, both the L- and K-function are less
    what is expected under the Poisson process the data. Thus they indicate a
    comparatively regular point pattern. Similarly, the pair-correlation
    function also suggests this.

### Exercise 2

The command `rThomas` generates simulated realisations of
the Thomas model (‘modified Thomas cluster process’).

1.  Read the help file.

    See `help("rThomas")`.

2.  Type `plot(rThomas(10, 0.05, 8))` a few times, and
    interpret the results.

    ```{r results="hide"}
    replicate(3, plot(rThomas(10, 0.05, 8), main = ""))
    ```

    A clustered process -- on average 10 clusters with 8 points. The standard
    deviation of the cluster distribution is 0.05, so most points will be within
    distance 0.10 from their parent.

3.  Experiment with the arguments of `rThomas` to obtain
    point patterns that
    
    i.  consist of a few, well-separated, very tight clusters of
        points;
        
    ii. look similar to realisations of a uniform Poisson process.

    We get few clusters by reducing the intensity of the parent process (first argument).
    Tightly and separated clusters are obtained by reducing the standard deviation
    (second argument).

    ```{r}
    plot(rThomas(5, scale = 0.01, mu = 8), main = "")
    ```

    If the are many clusters with a large standard deviation it looks like Poisson.

    ```{r}
    plot(rThomas(100, scale = 1, mu = 1), main = "")
    ```

### Exercise 3

1.  Read the help file for `kppm`.

    See `help("kppm")`.

2.  Fit the Thomas model to the `redwood` data by the
    method of minimum contrast:
    ```{r, eval=FALSE}
    fit <- kppm(redwood ~ 1, clusters="Thomas")
    fit
    plot(fit)
    ```

    From the documentation, the minmum contrast fitting procedure is default.
    Hence, we need not specify it.

    ```{r}
    fit <- kppm(redwood ~ 1, clusters = "Thomas")
    fit
    plot(fit, main = "", pause = FALSE) 
    ```

3.  Read off the parameters of the fitted model, and generate a
    simulated realisation of the fitted model using `rThomas`.

    From the previous output, we can read off the parameters to do the simulation
    (or we can use `parameters` to extract them):

    ```{r}
    (p <- parameters(fit))
    rt2 <- rThomas(kappa = p$kappa, scale = p$scale, mu = p$mu)
    plot(rt2, main = "")
    ```

4.  Type `plot(simulate(fit))` to generate a simulated realisation
    of the fitted model automatically.

    OK, let try that alternative:
    
    ```{r}
    plot(simulate(fit, drop = TRUE), main = "")
    ```

5.  Try the command
    ```{r eval=FALSE}
    fit2 <- kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=10, scale=0.1))
    ```
    and briefly explore the fitting algorithm’s sensitivity to the
    initial guesses at the parameter values `kappa` and `scale`.

    For "large" kappa (parent intensity) and "small" scale (standard deviation), the 
    algorithm seems quite robust:
    ```{r}
    kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=10, scale=0.1))
    kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=100, scale=0.01))
    ```

    However, for a very small parent intensity (kappa) and large offspring scale the
    fit changes considerably.
    
    ```{r}
    kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=0.1, scale=10))
    ```

6.  Generate and plot several simulated realisations of the fitted
    model, to assess whether it is plausible.

    ```{r}
    XX <- simulate(fit, nsim = 11)
    XX[[12]] <- redwood
    plot(XX, main = "", main.panel = "")
    ```
    
    The actual data do not look too different from the simulated (apart from the
    artificial discretisation in the real data which can be seen on larger plots).

7.  Extract and plot the fitted pair correlation function by
    ```{r, eval = FALSE}
    pcffit <- pcfmodel(fit)
    plot(pcffit, xlim = c(0, 0.3))
    ```

    OK, let's try that:
    
    ```{r}
    pcffit <- pcfmodel(fit)
    plot(pcffit, xlim = c(0, 0.3), main = "pair correlation")
    ```

8.  Type `plot(envelope(fit, Lest, nsim=39))` to generate simulation
    envelopes of the $L$ function from this fitted model. Do they
    suggest the model is plausible?

    ```{r}
    plot(envelope(fit, Lest, nsim = 39, global = TRUE))
    ```
    
    Yes, the model seems plausible and it does not devivate from the envelope.


### Exercise 4

1.  Fit a Matern cluster process to the `redwood` data.

    We fit the Matern cluster process by specifying the `clusters` argument to
    be `MatClust`.
    
    ```{r}
    mfit <- kppm(redwood ~ 1, clusters = "MatClust")
    ```

2.  Use `vcov` to estimate the covariance matrix of the parameter estimates.

    The variance (covariance matrix) is computed straightforwardly:

    ```{r}
    vcov(mfit)
    ```

3.  Compare with the covariance matrix obtained when fitting a homogeneous Poisson model.

    ```{r}
    vcov(ppm(redwood ~ 1))
    ```

    As can be seen, the variance of the intensity estimate is quite a bit larger
    in the Matern model. This comes naturally by the doubly stochastic
    construction of the Matern model.
