---
title: 'Lab 2: Poisson models -- solutions'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

```{r, message=FALSE}
library(spatstat)
```


### Exercise 1

The command `rpoispp(100)` generates realisations of the
Poisson process with intensity $\lambda = 100$ in the unit square.

1.  Repeat the command `plot(rpoispp(100))` several times
    to build your intuition about the appearance of a completely
    random pattern of points.

    Let's plot it three times:
    
    ```{r, results='hide'}
    replicate(3, plot(rpoispp(lambda = 100), main = ""))
    ``` 

    As can be seen, the points (unsurprisingly) are much more random that want one 
    might think. "Randomly" drawing points on a piece of paper one would usually
    draw a point pattern that is more regular (i.e. the points are repulsive).

2.  Try the same thing with intensity $\lambda = 1.5$.

    For brevity we only do it once here:
    
    ```{r, results='hide'}
    plot(rpoispp(lambda = 1.5), main = "")
    ``` 

    Here we expect 1.5 points in the plot each time.


### Exercise 2

Returning to the Japanese Pines data,

1.  Fit the uniform Poisson point process model to the Japanese
    Pines data
    ```{r, eval = FALSE}
    ppm(japanesepines~1)
    ```

    We fit the Poisson process model with the given command and print the output:
    
    ```{r}
    m.jp <- ppm(japanesepines ~ 1)
    print(m.jp)
    ```

2.  Read off the fitted intensity. Check that this is the correct
    value of the maximum likelihood estimate of the intensity.

    We extract the coeficient with the `coef` function, and compare to the
    straightforward estimate obtained by `intensity``:

    ```{r}
    unname(exp(coef(m.jp)))
    intensity(japanesepines)
    ```

    As seen, they agree exactly.

### Exercise 3

The `japanesepines` dataset is believed to exhibit
spatial inhomogeneity.

1.  Plot a kernel smoothed intensity estimate.

    Plot the kernel smoothed intensity estimate selecting the bandwidth with `bw.scott`:
    
    ```{r}
    jp.dens <- density(japanesepines, sigma = bw.scott)
    plot(jp.dens)
    plot(japanesepines, col = "white", cex = .4, pch = 16, add = TRUE)
    ```

2.  Fit the Poisson point process models with loglinear intensity
    (trend formula `~x+y`) and log-quadratic intensity (trend
    formula `~polynom(x,y,2)`) to the Japanese Pines data.

    We fit the two models with `ppm`:
    
    ```{r}
    jp.m <- ppm(japanesepines ~ x + y)
    jp.m2 <- ppm(japanesepines ~ polynom(x, y, 2) )
    ```

3.  extract the fitted coefficients for these models using
    `coef`.
        
    ```{r}
    coef(jp.m)
    coef(jp.m2)
    ```
    
4.  Plot the fitted model intensity (using `plot(predict(fit))`)

    ```{r}
    par(mar=rep(0,4))
    plot(predict(jp.m), main = "")
    plot(predict(jp.m, se=TRUE)$se, main = "")
    
    plot(predict(jp.m2), main = "")
    plot(predict(jp.m2, se=TRUE)$se, main = "")
    ```

5.  perform the Likelihood Ratio Test for the null hypothesis of a
    loglinear intensity against the alternative of a log-quadratic
    intensity, using `anova`.

    ```{r}
    anova(jp.m, jp.m2)
    ```

6.  Generate 10 simulated realisations of the fitted log-quadratic
    model, and plot them, using `plot(simulate(fit, nsim=10))` where 
    `fit` is the fitted model.

    ```{r}
    par(mar=rep(0.5,4))
    plot(simulate(jp.m2, nsim=10), main = "")
    ```

### Exercise 4

The `bei` dataset gives the locations of trees in a survey area
with additional covariate information in a list `bei.extra`.

1.  Fit a Poisson point process model to the data which assumes
    that the intensity is a loglinear function of terrain slope
    and elevation (hint: use `data = bei.extra` in `ppm`).

    We fit the log-linear intensity model with the following:
    
    ```{r}
    bei.m <- ppm(bei ~ elev + grad, data = bei.extra)
    ```

2.  Read off the fitted coefficients and write down the fitted
    intensity function.

    The coefficents are extraced with `coef`:
    
    ```{r}
    coef(bei.m)
    ```

    Hence the model is $log\lambda(u) = -8.55 + 0.02\cdot E(u) + 5.84 G(u)$ where
    $E(u)$ and $G(u)$ is the elevation and gradient, respectively, at $u$.

3.  Plot the fitted intensity as a colour image.

    ```{r}
    plot(predict(bei.m), main = "")
    plot(bei, cex = 0.3, pch = 16, cols = "white", add = TRUE)
    ```

4.  extract the estimated variance-covariance matrix of the
    coefficient estimates, using `vcov`.

    We call `vcov` on the fitted model object:
    ```{r}
    vcov(bei.m)
    ```

5.  Compute and plot the standard error of the intensity estimate
    (see `help(predict.ppm)`).

    From the documentation the argument `se` will trigger the computation of the 
    standard errors. These are then plotted in the standard manner.
    
    ```{r}
    std.err <- predict(bei.m, se = TRUE)$se
    plot(std.err, main = "")
    ```

### Exercise 5

The dataset `ants` is a multitype point pattern representing the
locations of nests of two species of ants.

1.  plot the data.

    ```{r}
    plot(ants)
    ```

2.  Fit the model `ppm(ants ~ marks)` and interpret the result.
    Compare the result with `summary(ants)` and explain
    the similarities.

    ```{r}
    fit1 <- ppm(ants ~ marks)
    ```
    
    This is a Poisson model with a separate constant intensity for each mark. The
    fitted intensities are:
    ```{r}
    exp(coef(fit1)[1])
    exp(coef(fit1)[1] + coef(fit1)[2])
    ```

    This agrees perfectly with the output of `summary(ants)`:
    ```{r}
    summary(ants)
    ```


3.  Fit the model `ppm(ants ~ marks + x)` and write down an
    expression for the fitted intensity function.

    ```{r}
    fit2 <- ppm(ants ~ marks + x)
    (co <- coef(fit2))
    ```
    
    Intensity for the reference type (Cataglyphis):
    
    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["x"]]` \cdot x)$$

    Intensity for the other type (Messor):

    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["marksMessor"]]` + `r co[["x"]]` \cdot x)$$


4.  Fit the model `ppm(ants ~ marks * x)` and write down an
    expression for the fitted intensity function.

    ```{r}
    fit3 <- ppm(ants ~ marks * x)
    (co <- coef(fit3))
    ```
    
    Intensity for the reference type (Cataglyphis):
    
    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["x"]]` \cdot x)$$
    
    Intensity for the other type (Messor):

    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["marksMessor"]]` + (`r co[["x"]]` + `r co[["marksMessor"]]`) \cdot x)$$

5.  Compute the fitted intensities of the three models fitted above 
    using `predict` and plot the results.

    ```{r}
    pred <- c(predict(fit1), predict(fit2), predict(fit3))
    plot(as.solist(pred), ncols = 2, main = "")
    ```

6.  Explain the difference between the models fitted by
    `ppm(ants ~ marks + x)` and `ppm(ants ~ marks * x)` .

    For the additive model the effect of the `x` coordinate is the same for both types
    of ants, while the effect of `x` differs in the multiplicative model.

## Exercise 6

The study region for the ants’ nests data `ants` is
divided into areas of "scrub" and "field". We want to fit a Poisson
model with different intensities in the field and scrub areas.

The coordinates of two points on the boundary line between field and
scrub are given in `ants.extra$fieldscrub`. First construct a function that
determines which side of the line we are on:

```{r}
fs <- function(x,y) {
  ends <- ants.extra$fieldscrub
  angle <- atan(diff(ends$y)/diff(ends$x))
  normal <- angle + pi/2
  project <- (x - ends$x[1]) * cos(normal) + (y - ends$y[1]) * sin(normal)
  factor(ifelse(project > 0, "scrub", "field"))
}
```

Now fit the models:
```{r, eval=FALSE}
ppm(ants ~ marks + side, data = list(side=fs))
ppm(ants ~ marks * side, data = list(side=fs))
```
and interpret the results.

```{r}
fit1 <- ppm(ants ~ marks + side, data = list(side=fs))
fit2 <- ppm(ants ~ marks * side, data = list(side=fs))
```

In the first model the fitted intensity is lower in the scrub than in the field
(but this effect is not significant).

In the second model the fitted intensity of Cataglyphis is lower in the scrub
than the intensity of Cataglyphis in the field, where as it is the other way
around for Messor. When we allow for the different effect between ant types the
scrub/field covariate is significant.

### Exercise 7

The `update` command can be used to re-fit a point
process model using a different model formula.

1.  Type the following commands and interpret the results:
    ```{r}
    fit0 <- ppm(japanesepines ~ 1)
    fit1 <- update(fit0, . ~ x)
    fit1
    fit2 <- update(fit1, . ~ . + y)
    fit2
    ```

    OK, let's do that:
    
    ```{r}
    fit0 <- ppm(japanesepines ~ 1)
    fit1 <- update(fit0, . ~ x)
    fit1
    fit2 <- update(fit1, . ~ . + y)
    fit2
    ```

2.  Now type `step(fit2)` and interpret the results.

    The backwards selection is done with the code:
    
    ```{r}
    step(fit2)
    ```
    
    First, given two models the preferred model is the one with the minimum AIC 
    value. In step 1, the removal of x results in the least AIC and is hence 
    deleted. In step 2, removing y results in a lower AIC than not deleing anything 
    and is thus deleted. This results in the constant model.
    
### Exercise 8

Fit Poisson point process models to the Japanese Pines data, with
the following trend formulas. Read off an expression for the fitted
intensity function in each case.

|Trend formula     |Fitted intensity function                   |
|:-----------------|:-------------------------------------------|
|`~1`              | $\log\lambda(u) = 4.17$                    |
|`~x`              | $\log\lambda(u) = 4.28 - 0.23x$            |
|`~sin(x)`         | $\log\lambda(u) = 4.29 - 0.26\sin(x)$      |
|`~x+y`            | $\log\lambda(u) = 4.07 - 0.23x + 0.42y$    | 
|`~polynom(x,y,2)` | $\log\lambda(u) = 4.06 + 1.14x - 1.56y - 0.75x^2 - 1.20xy + 2.51y^2$ |
|`~factor(x < 0.4)`| $\log\lambda(u) = 4.10 + 0.16\cdot I(x < 0.4)$ |

(Here, $I(\cdot)$ denote the indicator function.)

The fitted intensity functions have been written into the table based on the follwing model fits:

```{r}
coef(ppm1 <- ppm(japanesepines ~ 1)) 
coef(ppm2 <- ppm(japanesepines ~ x))
coef(ppm3 <- ppm(japanesepines ~ sin(x)))
coef(ppm4 <- ppm(japanesepines ~ x + y))
coef(ppm5 <- ppm(japanesepines ~ polynom(x, y, 2)))
coef(ppm6 <- ppm(japanesepines ~ factor(x < 0.4)))
```

### Exercise 9

Make image plots of the fitted intensities for the inhomogeneous
models above.

Again, we use `plot(predict())`:

```{r}
plot(predict(ppm1), main = "")
plot(predict(ppm2), main = "")
plot(predict(ppm3), main = "")
plot(predict(ppm4), main = "")
plot(predict(ppm5), main = "")
plot(predict(ppm6), main = "")
```

